\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-25 paper.tex
Tweaks

\newcommand{\noderef}[1]{\textsf{#1}}
\newcommand{\tsinfer}[0]{\texttt{tsinfer}}
\newcommand{\kwarg}[0]{\texttt{KwARG}}
\newcommand{\argweaver}[0]{\texttt{ARGweaver}}
\newcommand{\relate}[0]{\texttt{Relate}}
\newcommand{\espalier}[0]{\texttt{Espalier}}

@@

A genome ARG (gARG) is a directed acyclic graph in which nodes represent
haploid genomes and edges represent
genetic inheritance between an ancestor and a descendant.
The topology of a gARG specifies that genetic inheritance
occurred between particular
ancestors and descendants, but the graph connectivity
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination
(and other processes that lead to the partial genome inheritance)
 we ``annotate'' the edges with the genome
coordinates over which inheritance occured.
We can define a gARG formally as follows.
Let $N \subset \mathbb{N}$ be the set of nodes representing
the genomes in the gARG, $S \subseteq N$ be the set of sampled genomes.
% (but see Section~\ref{sec-retro-pro} for some subtlety around this point).
Then, $E$ is the set of edges, where each element
is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
parent nodes and $I$ is the set of disjoint genomic intervals
over which genome $c$ inherits from $p$.
Thus, each topological connection between
a parent and child node in the graph is annotated with a set of
inheritance intervals $I$.
Here, the terms parent and child are used in graph sense;
these nodes respectively represent ancestor and descendant genomes,
which can be separated by multiple generations.
We will use these two sets of terms interchangeably.

In many settings, nodes are dated, i.e.\ each
node $u\in N$ is associated with a time $\tau_u$.
The topological relationships
defined by the directed graph structure, however,
 is sufficient for many applications,
and we will assume that node dates are not known here.
% This should be obvious, but seems like it needs to be said
In practical settings, we will wish to associate additional
metadata with nodes such as sample identifiers or quality-control metrics.
It is therefore best to think of the
integers used here in the definition of a node as an \emph{identifier},
with which arbitrary additional information can be associated.
Concretely defining nodes as integers in this way has some
significant advantages. In particular, it provides a direct link
between an ARG and its local trees in terms of a well-defined
combinatorial object (Section~\ref{sec-ARG-and-local-trees}),
as well as enabling straightforward connection between
definitions and practical implementation (Section~\ref{sec-efficiency}).

@@

It is important that we make the distinction here between the local
trees that we can derive from a known ARG (as just discussed),
and the ARGs that we can derive from a sequence of \emph{estimated}
local trees.
The recent
\espalier~\citep{rasmussen2022espalier} ARG inference method
is illustrative in this context.
\espalier\ works by splitting an input sequence alignment into
segments that are assumed to be non-recombining. Within
each segment, an initial local tree is estimated using
standard phylogenetic methods. By necessity, these local
trees will contain internal nodes that are unlabelled and
consist only of branching points: there is no information
shared between the independent tree estimation steps
across segments. Part of the task of stitching
these trees together
into an ARG is then, essentially, to generate labels for
the internal nodes, and decide which nodes persist
across multiple local trees.
\espalier\ is specifically interested
in the details of recombination events,
and also attempts to infer
the precise subtree prune-and-regraft (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
induced by recombination between these partially reconciled local trees.
Inferring the SPRs between leaf-labelled trees is
NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but it is unclear what the complexity is when there
is a degree of internal node sharing between trees.
% I don't really know if this is true, it's just a way of wrapping the
% section up by saying that other things we thought were settled might
% be worth thinking about again.
The combinatorial formulation of ARGs and local trees provided here
may help clarify these fundamental questions.

@@

The \texttt{tskit} library is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with gARGs (encoded as a succinct tree sequence).
Based on core functionality written
in C, it provides interfaces in C, Python and Rust.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
adrion2020community,
terasaki2021geonomics,
baumdicker2021efficient,
fan2022genealogical,
guo2022recombination,
korfmann2022weak,
mahmoudi2022bayesian,
petr2022slendr,
rasmussen2022espalier,
zhang2023biobank,
nowbandegani2023extremely,
ignatieva2023distribution,
fan2023likelihood}.
The technical details of \texttt{tskit}, and how it provides an
efficient and portable platform for ARG-based analysis, are beyond
the scope of this manuscript.
In the interest of avoiding confusion, however, we list a
few minor details in which the formal details of gARGs
provided here differ from their practical implementation in
\texttt{tskit}.
Firstly, ``edges'' in tree sequence terminology would be perhaps better be
described as ``edge-intervals'', as each describes a single contiguous
interval of genome inheritance between a pair of nodes.
This denormalisation of the gARG data model in is for efficiency purposes.
Secondly, $-1$ is used to denote the presence of
roots in an oriented tree, rather than $0$ as used here for notational
simplicity.

>>>>

\newcommand{\noderef}[1]{\textsf{#1}}
\newcommand{\tsinfer}[0]{\texttt{tsinfer}}
\newcommand{\kwarg}[0]{\texttt{KwARG}}
\newcommand{\argweaver}[0]{\texttt{ARGweaver}}
\newcommand{\relate}[0]{\texttt{Relate}}
\newcommand{\espalier}[0]{\texttt{Espalier}}
\newcommand{\arbores}[0]{\texttt{Arbores}}

@@

A genome ARG (gARG) is a directed acyclic graph in which nodes represent
haploid genomes and edges represent
genetic inheritance between an ancestor and a descendant.
The topology of a gARG specifies that genetic inheritance
occurred between particular
ancestors and descendants, but the graph connectivity
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination
(and other processes that lead to partial genome inheritance)
 we ``annotate'' the edges with the genome
coordinates over which inheritance occured.
We can define a gARG formally as follows.
Let $N = \{1, \dots, n\}$ be the set of nodes representing
the genomes in the gARG, 
and  $S \subseteq N$ be the set of sampled genomes.
Then, $E$ is the set of edges, where each element
is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
parent nodes and $I$ is the set of disjoint genomic intervals
over which genome $c$ inherits from $p$.
Thus, each topological connection between
a parent and child node in the graph is annotated with a set of
inheritance intervals $I$.
Here, the terms parent and child are used in graph sense;
these nodes respectively represent ancestor and descendant genomes,
which can be separated by multiple generations.
We will use these two sets of terms interchangeably.

In many settings, nodes are dated, i.e.\ each
node $u\in N$ is associated with a time $\tau_u$.
The topological relationships
defined by the directed graph structure, however,
 is sufficient for many applications,
and we will assume that node dates are not known here.
% This should be obvious, but seems like it needs to be said
In practical settings, we will wish to associate additional
metadata with nodes such as sample identifiers or quality-control metrics.
It is therefore best to think of the
integers used here in the definition of a node as an \emph{identifier},
with which arbitrary additional information can be associated.
% We can delete then next two sentances if we want, but nodes are
% integers and this is a vital element of the model.
Concretely defining nodes as integers in this way has some
significant advantages. In particular, it provides a direct link
between an ARG and its local trees in terms of a well-defined
combinatorial object (Section~\ref{sec-ARG-and-local-trees}),
as well as enabling straightforward connection between
definitions and practical implementation (Section~\ref{sec-efficiency}).

@@

It is important that we make the distinction here between the local
trees that we can derive from a known ARG (as just discussed),
and the ARGs that we can derive from a sequence of \emph{estimated}
local trees.
The ARG inference methods \arbores~\citep{heine2018bridging} and
\espalier~\citep{rasmussen2022espalier} are illustrative in this context.
Both begin by splitting an input sequence alignment into
segments that are assumed to be non-recombining. Within
each segment, an initial local tree is estimated using
standard phylogenetic methods. By necessity, these local
trees will contain internal nodes that are unlabelled and
consist only of branching points: there is no information
shared between the independent tree estimation steps
across segments. Part of the task of stitching
these trees together
into an ARG is then, essentially, to generate labels for
the internal nodes, and decide which nodes persist
across multiple local trees. \arbores~bridges the gaps 
between the inferred local trees via a coalescent-based 
sampling scheme.
\espalier\ is specifically interested
in the details of recombination events,
and also attempts to infer
the precise subtree prune-and-regraft (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
induced by recombination between these partially reconciled local trees.
Inferring the SPRs between leaf-labelled trees is
NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but it is unclear what the complexity is when there
is a degree of internal node sharing between trees.
% I don't really know if this is true, it's just a way of wrapping the
% section up by saying that other things we thought were settled might
% be worth thinking about again.
The combinatorial formulation of ARGs and local trees provided here
may help clarify these fundamental questions.

@@

The \texttt{tskit} library is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with gARGs (encoded as a succinct tree sequence).
Based on core functionality written
in C, it provides interfaces in C, Python and Rust.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
adrion2020community,
terasaki2021geonomics,
baumdicker2021efficient,
fan2022genealogical,
guo2022recombination,
korfmann2022weak,
mahmoudi2022bayesian,
petr2022slendr,
rasmussen2022espalier,
zhang2023biobank,
nowbandegani2023extremely,
ignatieva2023distribution,
fan2023likelihood}.
The technical details of \texttt{tskit}, and how it provides an
efficient and portable platform for ARG-based analysis, are beyond
the scope of this manuscript.
In the interest of avoiding confusion, however, we list a
few minor details in which the formal details of gARGs
provided here differ from their practical implementation in
\texttt{tskit}.
Firstly, ``edges'' in tree sequence terminology would be perhaps better be
described as ``edge-intervals'', as each describes a single contiguous
interval of genome inheritance between a pair of nodes.
This denormalisation of the gARG data model is for efficiency purposes.
Secondly, $-1$ is used to denote the presence of
roots in an oriented tree, rather than $0$ as used here for notational
simplicity.
